A pipeline to analyse time-course gene expression data

### 2020-07-10 scripts/manuscript.Rmd
small change

---
title: "A pipeline to analyse time-course gene expression data"
author:
    - name: Nelle Varoquaux
      affiliation: CNRS, University Grenoble Alpes, TIMC-IMAG, Grenoble, France
    - name: Elizabeth Purdom
      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

- Make clear that it applies to RNASeq and how to call the different functions
    to RNASeq
- Check that clustering can deal with count data.
- Make sure package is compatible bioconductor

@@

# From Bioconductor
library(topGO)
library(biomaRt)
library(KEGGprofile)
library(BiocWorkflowTools)

@@

First let's load the data. We provide the normalized
data and metadata of [@shoemaker:ultrasensitive] for simplicity.

```{r}
# Now load in the metadata
meta = read.table("../data/shoemaker2015/meta.tsv")
data = read.table(gzfile("../data/shoemaker2015/data.tsv.gz"))
```

@@

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol; @park:evalution].

@@

```{r echo=FALSE}
contrasts[1:5]
```

@@

**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
**NV** Fair point… maybe this should be fixed somehow by adding a random
perturbation to the scores when the scores are equal? Although I think the
text is still appropriate: those genes should definitely be filtered out.

@@

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    moanin::rescale_values(data_to_plot)[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

@@

@@

```{r bootstrapped_clustering, eval=FALSE}
# You may want to set the random seed of the experiment so that the results
# don't vary if you rerun the experiment.
# set.seed(random_seed)
n_genes = dim(y)[1] * sample_proportion
indices = sample(1:dim(y)[1], n_genes, replace=TRUE)

@@

Now we will use the function `consensus_matrix` in the `moanin` package to
calculate proportion of times each pair of samples was clustered together
across the 25 resampling runs, and plot the heatmap of those consensus
matrices for $k=5$ and $k=20$.

@@

```{r echo=FALSE, results="hide"}
n_clusters = 2:20
all_labels = list()
for(i in n_clusters){
    filename = paste0("results/stability_", i, ".tsv")
    stability = read.table(filename, sep="\t")
    all_labels[[paste0("k=", i)]] = stability
}
```

@@

```{r plotModelExplorer, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k"}
plot_model_explorer(all_labels)
```

@@

```{r  plotModelExplorerZoom, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10"}
clusters = c("k=6", "k=7", "k=8", "k=9", "k=10")
selected_labels = all_labels[clusters]
plot_model_explorer(selected_labels)
```

@@

Once the gene ID to GO mapping list is created, `moanin` provides an interface
to `topGO` to determine enriched GO terms. In particular, it performs a
p-value correction and only returns the significant GO-term enrichment in an
easy to use `data.frame` object. Here, we show an example of running a GO term
enrichment on the "Biological process" ontology (BP).

>>>>

---
title: "A pipeline to analyse time-course gene expression data"
author:
    - name: Nelle Varoquaux
      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
    - name: Elizabeth Purdom
      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

- Make clear that it applies to RNASeq and how to call the different functions
    to RNASeq
- Check that clustering can deal with count data.
- Update Nelle's affiliation!

@@

# From Bioconductor
library(topGO)
library(biomaRt)

@@

First let's load the data. The package `moanin` contains the normalized
data and metadata of [@shoemaker:ultrasensitive].

```{r}
# Now load in the metadata
data(shoemaker2015)
meta = shoemaker2015$meta
data = shoemaker2015$data
```

@@

The first steps of analysis of gene expression data is always to do
normalization and quality control checks of the data. However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file. **EAP**: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.

@@

```{r echo=FALSE}
contrasts[1:5]
```
**EAP**: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like. I like to reserve table format for actual results, rather than code-like input

@@

**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.

@@

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    moanin:::rescale_values(data_to_plot)[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.

**EAPL** actually, why do we need to rescale them between 0-1 rather than just a normal call to `scale`? Would it really make a difference?

@@

**EAP** This didn't make sense to me: "run the clustering for all possible clusters of interest." Do you mean across different values of $k$? That's what I changed it to.

@@

**EAP**: Why don't you use the same rule/code as above? i.e. with the fisher p-values etc.? A little strange to have two different filtering rules for the clustering. Or is this the same, and you're restating the rule? (Unclear why you'd have to redo everything here...didn't go back and compare code)

```{r bootstrapped_clustering, eval=FALSE}
pvalues = de_analysis[, pval_col_to_keep]

splines_model = moanin::create_moanin_model(meta, degrees_of_freedom=6)
fishers_pval = moanin:::pvalues_fisher_method(pvalues)
fishers_qval = stats::p.adjust(fishers_pval)

# Keep only the genes found differentially expressed after combining the
# p-values using Fisher's method and with a log fold change greater than 2
genes_to_keep = row.names(
    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
y = as.matrix(data[genes_to_keep, ])

# You may want to set the random seed of the experiment so that the results
# don't vary if you rerun the experiment.
# set.seed(random_seed)
n_genes = dim(y)[1] * sample_proportion
indices = sample(1:dim(y)[1], n_genes, replace=TRUE)

@@

Now we will use the function `consensus_matrix` in the `moanin` package to
calculate proportion of times each pair of samples was clustered together
across the 20 resampling runs, and plot the heatmap of those consensus
matrices for $k=5$ and $k=20$.

**EAP** It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?

@@

```{r echo=FALSE, results="hide"}
n_clusters = 2:20
all_labels = list()
for(i in n_clusters){
    filename = paste0("results/stability_", i, ".tsv")
    stability = read.table(filename, sep="\t")
    all_labels[[paste0("B", i)]] = stability
}
```

@@

```{r plotModelExplorer, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k"}
plot_model_explorer(all_labels)
```
**EAP** I'm confused by this plot. Isn't the numbering here for different values of $k$? Why do the labels say "B1" and "B2"? Isn't that the notation above for the bootstrap replications for each values of $k$? It seems like the names of `all_labels` should be something like `k=1`,..., while the names of `all_labels[[1]]` should be `B1,B2,...`. But currently they have the same values.

@@

```{r  plotModelExplorerZoom, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10"}
clusters = c("B6", "B7", "B8", "B9", "B10")
selected_labels = all_labels[clusters]
plot_model_explorer(selected_labels)
```

@@

Once the gene ID to GO mapping list is created, `moanin` provides an 
interface to `topGO` to determine enriched GO terms. **EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe** Here, we show an example of
running a GO term enrichment on the "Biological process" ontology (BP).
